Resolvable Ambiguity

### 2019-05-30 paper/paper.tex
EOD

\newcommand{\NT}{\mathbb{N}} % Set of nonterminals
\newcommand{\T}{\Sigma} % Set of terminals
\newcommand{\Labels}{\mathbb{L}} % Set of labels
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{semantic}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}

@@

\begin{quote} % TODO: remove
  \textbf{NOTE:} The text before this section is largely out of date.
\end{quote}
% TODO: note that the syntax tree definition is a little bit non-standard

\subsection{Context-Free Grammars}

@@

\subsection{Syntax Trees}

A syntax tree takes the terminals of a word and arranges them into a tree, the shape of which is determined by how (i.e., using which productions) the word was parsed / recognized. For this paper we will use a linear encoding of syntax trees, letting us use context-free grammars to express the set of syntax trees for some given word language. For example, the parse trees of a context-free grammar $G_w = (\NT, \T, P, S)$ are described by the context-free grammar $G_t = (\NT, \T \cup \T_{[]}, P', S)$ where:

\begin{itemize}
\item $T_{[]} = \bigcup_{p \in P} \{ [_p, ]_p \}$, i.e., a unique pair of brackets per production.
\item $P' = \{ N -> [_p \alpha ]_p \mid p \in P \land p = N -> \alpha \}$, i.e., each production is wrapped with a unique pair of brackets.
\end{itemize}

\noindent Such a grammar will (slightly erroneously) be referred to as a tree grammar, and the recognized language as a tree language. We will also use syntax trees that are not directly derived from the parse trees of a context-free grammar. The precise construction of these will be described at the point of their introduction, but the basic process is the same; surround each production with a unique pair of brackets.

% TODO: reference a book on tree automata and languages
As such we can define a function $\yield$ (as seen, e.g., in ) that takes a tree and flattens it, producing the word it was constructed from. Intuitively, this function merely removes the added brackets.

\subsection{Ambiguity}

The standard definition of ambiguity, using the above formulation of syntax trees, is as follows:

\begin{definition}
A word $w \in L(G_w)$ is ambiguous relative its parse tree $G_t$ if:
  \[ \exists t_1, t_2 \in L(G_t).\ t_1 \neq t_2 \land \yield(t_1) = w = \yield(t_2)\]
\end{definition}

\begin{definition}
  A word $w \in L(G_w)$ is ambiguous relative a tree grammar $G_t$ and a function $\parse : L(G_w) -> 2^{L(G_t)}$ if:
  \[ |\parse(w)| > 1 \]
\end{definition}

\[ \parse(w) = \{ t \mid w = \yield(t) \land t \in L(G_t) \}\ \]

\subsection{Automata}

@@

\begin{itemize}
\item A finite set of states $Q$.
\item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
\item A finite set of stack symbols $\Gamma$, i.e., a stack alphabet.
\item A transition function $\delta: Q \times \Sigma \times \Gamma -> 2^{Q \times \Gamma^{*}}$.
\item An initial state $q_0 \in Q$.
\item A set of final states $F \subseteq Q$.
\end{itemize}

@@

\subsection{Visibly Pushdown Languages}

@@

@@

\begin{figure}
  \begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
    $E$ & $l$ & \verb|'['| ($E$ (\verb|';'| $E$)$^{*}$ $+$ $\epsilon$) \verb|']'| \\
    $E$ & $a$ & $E$ \verb|'+'| $E$ \\
    $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
    $E$ & $n$ & $N$ \\
  \end{tabular}
  \caption{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity. Assumes that $N$ matches a numeric terminal.}
  \label{fig:running-example-definition}
\end{figure}

@@

From $D$ we then generate four grammars: $G_w$, $G_t$, $G'_w$, and $G'_t$. The first two represent the \emph{what}, while the latter two represent the \emph{how}:

\begin{itemize}
\item $G_w$ represents a word language describing all semantically distinct programs.
\item $G_t$ represents a tree language describing the syntax trees obtained by parsing a word in $L(G_w)$. Note that the added brackets are not present in the productions of $E_{l1}$ and $E_{l2}$, since those are all part of the same labelled production in $D$ in Figure~\ref{fig:running-example-definition}.
\item $G'_w$ is essentially a modified version of $G_w$, e.g., adding parentheses and other forms of disambiguation.
\item $G'_t$ represents a tree language describing the syntax trees obtained by parsing a word in $L(G'_w)$.
\end{itemize}

\begin{figure}
  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      \toprule
      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E$ & $E$ \verb|'+'| $E$ \\
      $E$ & $E$ \verb|'*'| $E$ \\
      $E$ & $N$ \\
      \midrule
      $E_{l1}$ & $\epsilon$ \\
      $E_{l1}$ & $E$ $E_{l2}$ \\
      \midrule
      $E_{l2}$ & $\epsilon$ \\
      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
      \bottomrule
    \end{tabular}
    \caption{$G_w$, the generated abstract syntax.}
    \label{fig:running-example-generated:w}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.54\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
      \toprule
      $E$ & $[_{l}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l}$ \\
      $E$ & $[_{a}$ & $E$ \verb|'+'| $E$ & $]_{a}$ \\
      $E$ & $[_{m}$ & $E$ \verb|'*'| $E$ & $]_{m}$ \\
      $E$ & $[_{n}$ & $N$ & $]_{n}$ \\
      \midrule
      $E_{l1}$ & & $\epsilon$ & \\
      $E_{l1}$ & & $E$ $E_{l2}$ & \\
      \midrule
      $E_{l2}$ & & $\epsilon$ & \\
      $E_{l2}$ & & \verb|';'| $E$ $E_{l2}$ & \\
      \bottomrule
    \end{tabular}
    \caption{$G_t$, a language describing a linear encoding of the parse trees of $G_w$.}
    \label{fig:running-example-generated:t}
  \end{subfigure}

  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      \toprule
      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E$ & $E$ \verb|'+'| $E$ \\
      $E$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
      $E$ & $N$ \\
      $E$ & \verb|'('| $E$ \verb|')'| \\
      \midrule
      $E_{\{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E_{\{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
      $E_{\{a\}}$ & $N$ \\
      $E_{\{a\}}$ & \verb|'('| $E$ \verb|')'| \\
      \midrule
      $E_{l1}$ & $\epsilon$ \\
      $E_{l1}$ & $E$ $E_{l2}$ \\
      \midrule
      $E_{l2}$ & $\epsilon$ \\
      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
      \bottomrule
    \end{tabular}
    \caption{$G'_w$, the generated concrete syntax.}
    \label{fig:running-example-generated:w-prime}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.54\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
      \toprule
      $E$ & $[_{l, \emptyset}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l, \emptyset}$ \\
      $E$ & $[_{a, \emptyset}$ & $E$ \verb|'+'| $E$ & $]_{a, \emptyset}$ \\
      $E$ & $[_{m, \emptyset}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ & $]_{m, \emptyset}$ \\
      $E$ & $[_{n, \emptyset}$ & $N$ & $]_{n, \emptyset}$ \\
      $E$ & $[_{g, \emptyset}$ & \verb|'('| $E$ \verb|')'| & $]_{g, \emptyset}$ \\
      \midrule
      $E_{\{a\}}$ & $[_{l, \{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l, \{a\}}$ \\
      $E_{\{a\}}$ & $[_{m, \{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ & $]_{m, \{a\}}$ \\
      $E_{\{a\}}$ & $[_{n, \{a\}}$ & $N$ & $]_{n, \{a\}}$ \\
      $E_{\{a\}}$ & $[_{g, \{a\}}$ & \verb|'('| $E$ \verb|')'| & $]_{g, \{a\}}$ \\
      \midrule
      $E_{l1}$ & & $\epsilon$ & \\
      $E_{l1}$ & & $E$ $E_{l2}$ & \\
      \midrule
      $E_{l2}$ & & $\epsilon$ & \\
      $E_{l2}$ & & \verb|';'| $E$ $E_{l2}$ & \\
      \bottomrule
    \end{tabular}
    \caption{$G'_t$, a language describing a linear encoding of the parse trees of $G'_w$.}
    \label{fig:running-example-generated:t-prime}
  \end{subfigure}%
  \caption{The generated grammars.}
  \label{fig:running-example-generated}
\end{figure}

@@

\begin{figure}
  \begin{subfigure}[b]{.40\linewidth}
    \[1 + 2 * 3\]
    \caption{Example word in $L(G_w)$.}
  \end{subfigure}
  \begin{subfigure}[b]{.55\linewidth}
    \begin{center}
    \Tree [.{$m$}
        [.{$a$}
          [.{$n$} 1 ]
          +
          [.{$n$} 2 ] ]
        *
        [.{$n$} 3 ] ]
    \end{center}
    \[[_m [_a [_n \terminal{1} ]_n \terminal{+} [_n \terminal{2} ]_n ]_a \terminal{*} [_n \terminal{3} ]_n ]_m\]
    \caption{Example word in $L(G_T)$ and visualized tree.}
  \end{subfigure}

  \begin{subfigure}[b]{.40\linewidth}
    \[(1 + 2) * 3\]
    \caption{Example word in $L(G'_w)$.}
  \end{subfigure}
  \begin{subfigure}[b]{.55\linewidth}
    \begin{center}
    \Tree [.{$m, \emptyset$}
        [.{$g, \{a\}$}
          (
          [.{$a, \emptyset$}
            [.{$n, \emptyset$} 1 ]
            +
            [.{$n, \emptyset$} 2 ] ]
          ) ]
        *
        [.{$n, \{a\}$} 3 ] ]
    \end{center}
    \[[_{m} [_{g, \{a\}} \terminal{(} [_{a} [_{n} \terminal{1} ]_{n} \terminal{+} [_{n} \terminal{2} ]_{n} ]_{a} \terminal{)} ]_{g, \{a\}} \terminal{*} [_{n, \{a\}} \terminal{3} ]_{n, \{a\}} ]_{m}\]
    \caption{Example word in $L(G'_T)$ and visualized tree.}
  \end{subfigure}
  \caption{Example with words from each grammar that correspond to each other. We abbreviate $[_{n, \emptyset}$ as $[_n$. The non-bracket terminals in the tree languages appear underlined for increased clarity.}
  \label{fig:example-words-in-square}
\end{figure}
}

At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses. For example, $G_w$ in Figure~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions). However, starting with that language definition would allow the expression ''$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ''$;2$'' is not a valid expression.

As an example, in Figure~\ref{fig:example-grammar}, \ref{fig:example-grammar:ambig-grammar} is the \emph{what} and \ref{fig:example-grammar:unambig-grammar} is the \emph{how}. The latter grammar is a modification of the former that adds precedence, associativity, and parentheses, yielding an unambiguous grammar with at least one way to express each semantically distinct tree.

\begin{figure*}
  \begin{subfigure}{.45\linewidth}
    \setlength{\grammarindent}{5em}
    \begin{grammar}
      <expr> ::= Sum: <expr> '+' <expr>
        \alt Product: <expr> '*' <expr>
        \alt Number: <number>
    \end{grammar}
    \caption{The (ambiguous) intuitive grammar without parentheses.}
    \label{fig:example-grammar:ambig-grammar}
  \end{subfigure}
  \begin{subfigure}{0.45\linewidth}
    \setlength{\grammarindent}{5em}
    \begin{grammar}
      <expr> ::= Sum: <product> '+' <expr>
        \alt PassthroughE: <product>

      <product> ::= Product: <atom> '*' <product>
        \alt PassthroughP: <atom>

      <atom> ::= Paren: '(' <expr> ')'
        \alt Number: <number>
    \end{grammar}
    \caption{The (unambiguous) grammar with parentheses.}
    \label{fig:example-grammar:unambig-grammar}
  \end{subfigure}

  \begin{subfigure}{.55\linewidth}
    \begin{center}
    \Tree [.PassthroughE
      [.\underline{Product}
        [.Paren
          '('
          [.\underline{Sum}
            [.PassthroughP [.\underline{Number} 1 ] ]
            '+'
            [.PassthroughE [.PassthroughP [.\underline{Number} 2 ] ] ] ]
          ')' ]
        '*'
        [.PassthroughP [.\underline{Number} 3 ] ] ] ]
    \end{center}
    \caption{Parse tree for $(1 + 2) * 3$. The underlined nodes are semantically important.}
    \label{fig:example-grammar:tree}
  \end{subfigure}
  \begin{subfigure}{.35\linewidth}
    \begin{center}
    \Tree [.\underline{Product}
      [.\underline{Sum} [.\underline{Number} 1 ] '+' [.\underline{Number} 2 ] ]
      '*'
      [.\underline{Number} 3 ] ]
    \end{center}
    \caption{The same parse tree, after removing the semantically unimportant nodes.}
    \label{fig:example-grammar:sem-tree}
  \end{subfigure}

  \caption{A basic expression grammar in two variations, and two example syntax trees.}
  \label{fig:example-grammar}
\end{figure*}

Our definition thus refers to four grammars in total:

\begin{itemize}
  \item The semantic grammar $G$, containing only the semantically important productions (e.g., Figure~\ref{fig:example-grammar:ambig-grammar}).
  \item $T_G$ (generally abbreviated as $T$), the parse trees of $G$, representing the trees that must be expressible (Figure~\ref{fig:example-grammar:sem-tree} is in this language).
  \item The parse grammar $G'$, a modification of $G$ meant to actually be used for parsing (e.g., Figure~\ref{fig:example-grammar:unambig-grammar}). The identifiers in this grammar need not be unique, but should be a superset of the identifiers in $G$. The modifications we will consider in this paper are introduced in Section~\ref{sec:modifications}.
  \item $T_{G'}$ (generally abbreviated as $T'$), the parse trees of $G'$ (Figure~\ref{fig:example-grammar:tree} is in this language).
\end{itemize}

\noindent We also require a function $\semantic : L(T') -> L(T)$ that removes the semantically unimportant productions from a parse tree. The relation between the four grammars can be seen in Figure~\ref{fig:grammar-square}. Finally, we define the function $\parse : L(G') -> 2^{L(T)}$, and its inverse $\words : L(T) -> 2^{L(G')}$:

$$
\begin{array}{rcl}
\parse(w') & = & \{ \semantic(t') \mid t' \in L(T') \land \yield(t') = w' \} \\
\words(t) & = & \{ w' \mid t \in \parse(w') \} \\
\end{array}
$$

\noindent The latter will be useful later, while $\parse$ is used directly in the definition of resolvable ambiguity:

\begin{figure}
  \begin{tikzcd}
    L(G) & L(T) \arrow[l, "\yield"] \\
    L(G') & L(T') \arrow[u, "\semantic"'] \arrow[l, "\yield"]
  \end{tikzcd}
  \caption{The grammars considered, and their relation to each other. $G$ is provided by a user of the system, along with instructions how to modify $G$ to construct $G'$, while $T$ and $T'$ are automatically derived.}
  \label{fig:grammar-square}
\end{figure}

>>>>

\newcommand{\NT}{\mathbb{N}} % Set of nonterminals
\newcommand{\T}{\Sigma} % Set of terminals
\newcommand{\Labels}{\mathbb{L}} % Set of labels
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{unparen}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}

@@

This section briefly describes the theoretical foundations we build upon. Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation. Section~\ref{sec:preliminaries-vpls} then describes visibly pushdown languages, which enable the analyses described in Sections~\ref{sec:static} and \ref{sec:dynamic}. Finally, Sections~\ref{sec:preliminaries-trees} and \ref{sec:preliminaries-ambiguity} describe trees and ambiguity, the former with some non-standard notation and the latter with a slightly wider definition than normal.

\subsection{Context-Free Grammars} \label{sec:preliminaries-cfgs}

@@

\subsection{Automata} \label{sec:preliminaries-automata}

@@

\begin{itemize}
\item A finite set of states $Q$.
\item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
\item A finite set of stack symbols $\Gamma$, i.e., a stack alphabet, including a special bottom of stack symbol $\bot$.
\item A transition function $\delta: Q \times \Sigma \times \Gamma -> 2^{Q \times \Gamma^{*}}$.
\item An initial state $q_0 \in Q$.
\item A set of final states $F \subseteq Q$.
\end{itemize}

@@

\subsection{Visibly Pushdown Languages} \label{sec:preliminaries-vpls}

@@

\subsection{Unranked Regular Tree Grammars} \label{sec:preliminaries-trees}

Trees generalize words by allowing each terminal to have more than one successor. The successors have an associated ordering, i.e., in the presence of multiple successors, one is considered to be first, one second, etc. Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors. This is as opposed to \emph{unranked} tree languages, where the arity of a terminal is not fixed. In this case, the sequence of successors to a single terminal tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.

Our later results and properties are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature. We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves, to give a more natural definition of $\yield$.

An unranked tree grammar $T$ is a tuple $(\NT, \T, X, P, S)$ where:

\begin{itemize}
\item $\NT$ is a set of (zero-arity) non-terminals.
\item $\T$ is a set of zero-arity terminals, used as leaves.
\item $X$ is a set of terminals without fixed arity, used as inner nodes or leaves.
\item $P$ is a set of productions, where a production has the form $N -> x(r)$ for $N \in \NT$, $x \in X$, and $r$ a regular expression over $\NT$ and $\T$.
\end{itemize}

\noindent A tree $t$ is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$. The set of trees recognized by $T$ is written $L(T)$\footnote{All word grammars will be named $G$ (possibly with a subscript), while all tree grammars will be named $T$ (possibly with a subscript), lessening the risk of confusion between the two cases.}, and is said to be the tree language of $T$.

$\yield$ then, is the sequence $w \in \T^{*}$ of terminals $t \in \T$ obtained by a left-to-right traversal of a tree. Informally, it is the flattening of a tree after all internal nodes have been removed.

\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}

The standard definition of ambiguity, given a grammar $G$ and its parse trees $T_G$, is as follows:

\begin{definition}
A word $w \in L(G)$ is ambiguous if:
  \[ \exists t_1, t_2 \in L(T_G).\ t_1 \neq t_2 \land \yield(t_1) = w = \yield(t_2) \]
\end{definition}

\begin{definition}
  A word $w \in L(G)$ is ambiguous relative a tree grammar $T$ and a function $\parse : L(G) -> 2^{L(T)}$ if:
  \[ |\parse(w)| > 1 \]
\end{definition}

\[ \parse(w) = \{ t \mid w = \yield(t) \land t \in L(T_G) \}\ \]

@@

\begin{figure}
  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
    $E$ & $l$ & \verb|'['| ($E$ (\verb|';'| $E$)$^{*}$ $+$ $\epsilon$) \verb|']'| \\
    $E$ & $a$ & $E$ \verb|'+'| $E$ \\
    $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
    $E$ & $n$ & $N$ \\
  \end{tabular}
  \caption{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity. Assumes that $N$ matches a numeric terminal.}
  \label{fig:running-example-definition}
\end{figure}

@@

From $D$ we then generate four grammars: $G_D$, $T_D$, $G'_D$, and $T'_D$. The first two represent the \emph{what}, while the latter two represent the \emph{how}:

\begin{itemize}
\item $G_D$ represents a word language describing all semantically distinct programs.
\item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding parentheses and other forms of disambiguation.
\item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
\end{itemize}

\begin{figure}
  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      \toprule
      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E$ & $E$ \verb|'+'| $E$ \\
      $E$ & $E$ \verb|'*'| $E$ \\
      $E$ & $N$ \\
      \midrule
      $E_{l1}$ & $\epsilon$ \\
      $E_{l1}$ & $E$ $E_{l2}$ \\
      \midrule
      $E_{l2}$ & $\epsilon$ \\
      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
      \bottomrule
    \end{tabular}
    \caption{$G_D$, the generated abstract syntax.}
    \label{fig:running-example-generated:w}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.54\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T_D$, the parse trees of $G_D$.}
    \label{fig:running-example-generated:t}
  \end{subfigure}

  \begin{subfigure}[t]{.45\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
      \toprule
      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E$ & $E$ \verb|'+'| $E$ \\
      $E$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
      $E$ & $N$ \\
      $E$ & \verb|'('| $E$ \verb|')'| \\
      \midrule
      $E_{\{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| \\
      $E_{\{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
      $E_{\{a\}}$ & $N$ \\
      $E_{\{a\}}$ & \verb|'('| $E$ \verb|')'| \\
      \midrule
      $E_{l1}$ & $\epsilon$ \\
      $E_{l1}$ & $E$ $E_{l2}$ \\
      \midrule
      $E_{l2}$ & $\epsilon$ \\
      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
      \bottomrule
    \end{tabular}
    \caption{$G'_D$, the generated concrete syntax.}
    \label{fig:running-example-generated:w-prime}
  \end{subfigure}%
%
  \begin{subfigure}[t]{.54\linewidth}
    \centering
    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
      \toprule
      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E$ & $n$ & $N$ $)$ \\
      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \midrule
      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
      \bottomrule
    \end{tabular}
    \caption{$T'_D$, the parse trees of $G'_D$.}
    \label{fig:running-example-generated:t-prime}
  \end{subfigure}%
  \caption{The generated grammars.}
  \label{fig:running-example-generated}
\end{figure}

@@

\begin{figure}
  \begin{subfigure}[b]{.40\linewidth}
    \[1 + 2 * 3\]
    \caption{Example word in $L(G_D)$.}
  \end{subfigure}
  \begin{subfigure}[b]{.55\linewidth}
    \begin{center}
    \Tree [.{$m$}
        [.{$a$}
          [.{$n$} \terminal{1} ]
          \terminal{+}
          [.{$n$} \terminal{2} ] ]
        \terminal{*}
        [.{$n$} \terminal{3} ] ]
    \end{center}
    \[m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))\]
    \caption{Example tree in $L(T_D)$ and visualization.}
  \end{subfigure}

  \begin{subfigure}[b]{.40\linewidth}
    \[(1 + 2) * 3\]
    \caption{Example word in $L(G'_D)$.}
  \end{subfigure}
  \begin{subfigure}[b]{.55\linewidth}
    \begin{center}
    \Tree [.{$m$}
        [.{$g$}
          \terminal{(}
          [.{$a$}
            [.{$n$} \terminal{1} ]
            \terminal{+}
            [.{$n$} \terminal{2} ] ]
          \terminal{)} ]
        \terminal{*}
        [.{$n$} \terminal{3} ] ]
    \end{center}
    \[m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))\]
    \caption{Example tree in $L(T'_D)$ and visualization.}
  \end{subfigure}
  \caption{Example with elements from each generated language that correspond to each other. The leaf terminals in the tree languages appear underlined to distinguish the two kinds of parentheses.}
  \label{fig:example-words-in-square}
\end{figure}
}

At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses. For example, $G_D$ in Figure~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions). However, starting with that language definition would allow the expression ''$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ''$;2$'' is not a valid expression.

We also require a function $\semantic : L(T'_D) -> L(T_D)$ that removes the semantically unimportant productions from a parse tree, i.e., it replaces every subtree $g($ \verb|'('| $t$ \verb|')'| $)$ with $t$. The relation between the four grammars can be seen in Figure~\ref{fig:grammar-square}. Finally, we define the function $\parse : L(G'_D) -> 2^{L(T_D)}$, and its inverse $\words : L(T_D) -> 2^{L(G'_D)}$:

$$
\begin{array}{rcl}
\parse(w') & = & \{ \semantic(t') \mid t' \in L(T'_D) \land \yield(t') = w' \} \\
\words(t) & = & \{ w' \mid t \in \parse(w') \} \\
\end{array}
$$

\noindent The latter will be useful later, while $\parse$ is used directly in the definition of \emph{resolvable ambiguity}:

\begin{figure}
  \begin{tikzcd}
    L(G_D) & L(T_D) \arrow[l, "\yield"] \\
    L(G'_D) & L(T'_D) \arrow[u, "\semantic"'] \arrow[l, "\yield"]
  \end{tikzcd}
  \caption{The grammars considered, and their relation to each other.}
  \label{fig:grammar-square}
\end{figure}

\begin{definition}
  A word $w \in L(G'_D)$ defined by the language definition $D$ is resolvably ambiguous if:
  $$
  \begin{array}{l}
  \forall t \in \parse(w).\\
  \quad \exists w' \in L(G'_D).\ \parse(w') = \{t\}
  \end{array}
  $$
\end{definition}

\noindent As an example, in the language defined in Figure~\ref{fig:running-example-definition}, the word \verb|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where

\begin{center}
  \begin{tabular}{l}
    $t_1 = a($ \hphantom{$a($} $n($ \verb|'1'| $)$ \verb|'+'| $a($ $n($ \verb|'2'| $)$ \hphantom{$)$} \verb|'+'| $n($ \verb|'3'| $)$ $)$ $)$ \\
    $t_2 = a($ $a($ $n($ \verb|'1'| $)$ \verb|'+'| \hphantom{$a($} $n($ \verb|'2'| $)$ $)$ \verb|'+'| $n($ \verb|'3'| $)$ \hphantom{$)$} $)$ \\
  \end{tabular}
\end{center}

\noindent This ambiguity can be resolved however, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$. To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:

\begin{center}
  \begin{tabular}{l}
    $t_3 = l($ \verb|'['| \hphantom{$s($} $n($ \verb|'1'| $)$ \verb|';'| $n($ \verb|'2'| $)$ \hphantom{$)$} \verb|']'| $)$ \\
    $t_4 = l($ \verb|'['| $s($ $n($ \verb|'1'| $)$ \verb|';'| $n($ \verb|'2'| $)$ $)$ \verb|']'| $)$ \\
  \end{tabular}
\end{center}

\noindent In this case, $t_4$ has an unambigous word (namely \verb|'[(1 ; 2)]'|), but $t_3$ does not. The solution is to modify the language definition in Figure~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_1\}$.

We can expand this to an entire language, stating that a language is resolvably ambiguous iff all its words are resolvably ambiguous, alternatively:

\begin{definition}
  A language defined by the language definition $D$ is resolvably ambiguous if:
  $$
  \begin{array}{l}
  \forall t \in L(T_D).\\
  \quad \exists w' \in L(G'_D).\ \parse(w') = \{t\}
  \end{array}
  $$
\end{definition}

\noindent Intuitively, a language is resolvably ambiguous if there is at least one (unambiguous\footnote{Using $\parse$ and $T_D$, i.e., relating $G'_D$ and $T_D$, instead of $G'_D$ and its parse trees.}) word for each semantically distinct tree. Note that neither $G_D$ nor $G'_D$ necessarily need to be unambiguous for this to hold.

We can now state the two central problems we wish to consider:

\begin{itemize}
\item Static resolvability analysis: determine if a given language definition $D$ is resolvably ambiguous, otherwise produce an unresolvably ambiguous word $w \in L(G'_D)$. Section~\ref{sec:static} provides a partial solution to this problem.
\item Dynamic resolvability analysis: given a language definition $D$ and a word $w \in L(G'_D)$, determine if it is resolvably ambiguous, and if so, present an unambiguous word for every tree $t \in \parse(w)$. Section~\ref{sec:dynamic} provides a full solution with one caveat: it only considers languages with balanced parentheses.
\end{itemize}
